---
title: Households by AMI Forecast
format:
  html:
   code-fold: true
   code-summary: "Show the code"
   code-tools: true
   toc: true
   output: false
---

# Purpose

This script is to generate a forecast of households by AMI based on the most recent three year average at the state level.

```{r}
#| label: setup

library(tidyverse)
library(flexiblas)
library(sdotools)
library(sdodemog)
library(googledrive)
library(DBI)
library(RPostgres)

# Enable parallel processing with FlexiBLAS
flexiblas_load_backend("OPENBLAS-OPENMP")
flexiblas_set_num_threads(parallel::detectCores() - 1)

```

# Pull Data from Postgres

The necessary tables are stored in the SDO Postgres database.

```{r}
#| label: pull_data

con <- sdodemog::connect_to_sdo_db()

hh_by_ten_ami <- pull_db_table(
  schema = "econ",
  table = "cost_burden_by_ami"
)

hh_projections <- pull_db_table(
  schema = "estimates",
  table = "household_projections"
)

xwalk_hna_region <- pull_db_table(
  "xwalk",
  "region_hna"
) |>
  mutate(region = str_sub(region_code, -1))

```

# Process Data

Begin by calculating by year the percentage of all households by tenure, and percentage of households by AMI group within each tenure group.

```{r}
#| label: check_region_to_state_totals

state_region_check <- hh_by_ten_ami |>
  filter(year >= 2021) |>
  mutate(region_type = if_else(region == "0", "State", "Region")) |>
  summarize(
    across(
      c(cost_burdened_30, cost_burdened_50, total_hh),
      sum,
      na.rm = TRUE
    ),
    .by = c(
      model_id,
      year,
      # tenure,
      ami_group,
      region_type
    )
  ) |>
  pivot_wider(
    names_from = region_type,
    values_from = c(cost_burdened_30, cost_burdened_50, total_hh)
  )


```

```{r}
#| label: process_data

calc_ami_allocation <- function(df) {
  hh_ami_pct <- df |>
    filter(!(is.na(ami_group))) |>
    select(year, tenure, ami_group, region, hh_ami_ten = total_hh) |>
    mutate(hh_tot = sum(hh_ami_ten), .by = c(year, region)) |>
    mutate(
      hh_ten = sum(hh_ami_ten),
      .by = c(year, region, tenure)
    ) |>
    mutate(
      pct_hh_ten = hh_ten / hh_tot,
      pct_hh_ami_ten = hh_ami_ten / hh_ten
    )

  hh_ami_pct_fcst <- hh_ami_pct |>
    filter(year >= max(year) - 2) |>
    summarize(
      hh_ami_ten_avg = mean(hh_ami_ten),
      hh_tot_avg = mean(hh_tot),
      hh_ten_avg = mean(hh_ten),
      .by = c(region, tenure, ami_group)
    ) |>
    mutate(
      pct_hh_ten = hh_ten_avg / hh_tot_avg,
      pct_hh_ami_ten = hh_ami_ten_avg / hh_ten_avg
    )

  hh_ami_allocation_tbl <- bind_rows(
    hh_ami_pct |>
      select(year, region, tenure, ami_group, pct_hh_ten, pct_hh_ami_ten),
    expand_grid(
      year = c((max(hh_ami_pct$year) + 1):max(hh_projections$year)),
      hh_ami_pct_fcst |>
        select(region, tenure, ami_group, pct_hh_ten, pct_hh_ami_ten)
    )
  )

  return(
    list(
      "hh_ami_pct" = hh_ami_pct,
      "hh_ami_pct_fcst" = hh_ami_pct_fcst,
      "hh_ami_allocation_tbl" = hh_ami_allocation_tbl
    )
  )
}

hh_ami_tbls <- hh_by_ten_ami |>
  calc_ami_allocation()

```

```{r}

hh_ami_forecast_region <- hh_projections |>
  filter(
    area_code != 0,
    household_type_id == 0,
    age_group_id == 0
  ) |>
  left_join(
    xwalk_hna_region |>
      select(
        area_code = countyfips,
        region_code
      ) |>
      mutate(region = str_sub(region_code, -1)),
    by = "area_code"
  ) |>
  summarize(
    total_households = sum(total_households, na.rm = TRUE),
    .by = c(year, region)
  ) |>
  left_join(
    hh_ami_tbls$hh_ami_allocation_tbl,
    by = c("year", "region")
  ) |>
  mutate(
    hh_allocated = total_households * pct_hh_ten * pct_hh_ami_ten
  )

```

```{r}
hh_ami_forecast_state <- hh_ami_forecast_region |>
  summarize(
    region = "0",
    hh_allocated = sum(hh_allocated),
    .by = c(year, tenure, ami_group)
  ) |>
  mutate(
    total_households = sum(hh_allocated),
    .by = year
  ) |>
  mutate(
    pct_hh_ten = sum(hh_allocated) / total_households,
    pct_hh_ami_ten = hh_allocated / sum(hh_allocated),
    .by = c(year, tenure)
  )

hh_ami_forecast <- bind_rows(
  hh_ami_forecast_state,
  hh_ami_forecast_region
) |>
  arrange(year, region, tenure, ami_group)

# check allocations = total
hh_ami_forecast |>
  summarize(
    hh_allocated = sum(hh_allocated),
    .by = c(year, region, total_households)
  ) |>
  mutate(check = total_households - hh_allocated) |>
  filter(abs(check) > 1)

```

```{r}
#| label: write_data

sdodemog::sdo_db_write_table(
  connect_to_sdo_db(),
  schema = "econ",
  table = "hh_ami_forecast",
  df = hh_ami_forecast,
  vintage = 2024,
  version = "v3",
  overwrite = TRUE,
  use_model_tracking = TRUE,
  latest = TRUE
)


write_csv(
  hh_ami_forecast,
  "data/hh_ami_forecast.csv"
)

```